---
title: "DECO"
output:
  md_document:
    variant: markdown_github
    toc: true

---

# Introduction

$\texttt{DECO}$ jointly models de novo mutations (DNMs) of parent-offspring trios, rare case/control variants and gene-set information to prioritize risk genes for the tested disease.

# Usage

Users can go directly to [this step](#run1) to run $\texttt{DECO}$. However, some details are described here.

There are two ways to obtain analysis results of $\texttt{DECO}$:
  
  1) Running $\texttt{DECO}$ for both variant and gene-set data. It takes some minutes or some hours for ~ 20,000 genes to obtain all genetic and gene-set parameters.
  
  2) Using genetic parameters from previous studies or from 1) to run $\texttt{DECO}$. It takes approximately 3 seconds for one gene-set. This approach can be used to analyze some thousands of  gene-sets.
  

## Run $\texttt{DECO}$ for both variant and gene-set data.

Below is an example for jointly analyzing variant data of schizophrenia (SCZ) and a gene-set. The SCZ data is from [https://genomemedicine.biomedcentral.com/articles/10.1186/s13073-017-0497-y](Nguyen et al., 2017, Genome Medicine).


### Data format
The data should be formatted as the $\texttt{xInputGS}$ table below:

  - Mutation rates should start with $\texttt{mut_}$.
  
  - De novo mutations should start with $\texttt{dn_}$.
  
  - Case variant data should start with $\texttt{cc_case}$.
  
  - Control variant data should start with $\texttt{cc_control}$.
  
  - Gene-set data should be coded 1 (presence) or 0 (absence) and in the column $\texttt{GS}$.
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

```{r}
xInputGS <- read.table("data/extTADA_SCZ_constrainedGenes.csv", header = TRUE, as.is = TRUE)
head(xInputGS)
```

### Use $\texttt{DECO}$ function {#run1}

The only function to obtain final results is $\texttt{DECO}$.

```{r}
##SET PARAMETERS for DECO
nIteration = 1000 ###This should be higher (e.g, >=5,000)
nChain = 2
nCore = nChain
ntrio = c(1077, 1077, 1077) #Trio numbers: there are three categories of de novo mutations 
ncase = c(3157, 1091, 1353) #Case numbers: there are three population samples of rare case/control variants
ncontrol = c(4672, 1193, 4769) #Control numbers: there are three population samples of rare case/control variants

###RUN DECO
source('script/DECO.R') #Load the source code

###Obtain results
outSCZ <- DECO(
        inputData = xInputGS, ## Input data should be formatted as above
        Ndn = array(c(ntrio)), #rep(ntrio, 1), ##Two de novo categories
        Ncase = array(ncase), #rep(N$ca, 1), ##Two case categories
        Ncontrol = array(ncontrol), #rep(N$cn, 1), ##Two control categories
        nIteration = nIteration,# nIteration ## Number of iterations: should be upto higher for real data
        nChain = nChain, #Number of MCMC chains
        nCore = nCore #Number of computer cores
    )

```

### Outputs of $\texttt{DECO}$

Output is a list of:

  1) $\texttt{dataPP}$: all input data, posterior probabilities (PP) and FDRs of genes.
  
  2) $\texttt{pars}$: estimates of parameters.
  
  Note: $\pi$ is not estimated directly, but $\alpha$ was estimated with $\pi = e^\alpha/(e^{1 + \alpha})$.
  
  3) $\texttt{genesetInfo}$: p-value ($\texttt{pGS}$), $\texttt{q1}$ (q01) and $\texttt{q0}$ (q00) of the tested gene-set.
  
  4) $\texttt{mcmcData}$: MCMC results of parameters.


We will take a look at these output information.

```{r}
outSCZ$genesetInfo
fdr1 <- outSCZ$dataPP
dim(fdr1[fdr1$FDR < 0.05, ])
```

## Using genetic parameters from previous studies or from 1) to run $\texttt{DECO}$

We describe an example in which we used genetic parameters from 1) to run $\texttt{DECO}$. To use this approach, we need Bayes Factors (BF) and $\pi$.
```{r}
outSCZ$pars
```
  
#### Read a gene-set (e.g. pLI09 in this example)

```{r}
gsData <- readLines('data/pLI09.txt')
xInputGS2 <- fdr1[, 1:14]
gsInput <- rep(0, dim(xInputGS2)[1])
gsInput[is.element(xInputGS2[, 1], gsData)] <- 1

table(gsInput)

pi0 <- as.numeric(outSCZ$pars[1, 'Mode'])
xInputGS2[, 'GS'] <- gsInput
head(xInputGS2, 2)

```

#### Run $\texttt{DECO}$

```{r}
outSCZ2 <- DECO(
  inputData = xInputGS2, ## Input data should be formatted as above
  pi0 = pi0,
  Ndn = array(c(ntrio)), #rep(ntrio, 1), ##Two de novo categories
  Ncase = array(ncase), #rep(N$ca, 1), ##Two case categories
  Ncontrol = array(ncontrol)
    )

```

#### Results

```{r}
fdr2 <- outSCZ2$dataPP
fdr2[fdr2$FDR < 0.05, ]
outSCZ2$genesetInfo
```
```{r}
stan_trace(outSCZ$mcmcData)
```

# Citation
Tan-Hoang Nguyen, Xin He, Ruth Brown, Bradley T. Webb, Eli A. Stahl, Kenneth S. Kendler, Vladimirov Vladimir, Brien P. Riley, Silviu-Alin Bacanu. *DECO: Prioritizing risk genes for schizophrenia by jointly analyzing rare variants and biological pathways*. Under Review (2020).
